Effective Inclusion of Electronic Polarization to a Negatively Charged Membrane

### 2019-03-06 Manuscript/manuscript.tex
Update text

The interaction of various cations including calcium 
with  neutral and also negatively charged lipid bilayers
was thoroughly compared between simulations and experiments 
from the point of the electrometer concept \cite{roux90, seelig87, altenbach84}
in the references~\citenum{catte16, nmrlipids_proj4}. 
It was concluded that the interaction of calcium with zwitterionic PC bilayers 
is generally overestimated in all studied models \cite{catte16},
while none of the models was capable of interpreting the NMR structural parameters
with the negatively charged PC:PS (5:1) bilayers \cite{nmrlipids_proj4}. 
Accounting for the electronic polarization was identified as 
a non-negligible correction to the interaction between cations and POPC bilayers in Ref.~\cite{melcr18}. 
Here, we demonstrate that the same approach 
significantly improves the description of calcium binding 
also to the negatively charged PC:PS (5:1) bilayers. 
In correspondence with the Ref.~\citenum{nmrlipids_proj4},
we also use the changes of the head group order parameters $S^\alpha$ and $S^\beta$ 
in PC and PS lipids 
to quantify the binding affinity of the calcium cations
and the induced structural changes
(Fig.~\ref{fig:delta_ordPar_CaCl_PCPS}). 
To directly connect to the previous works \citep{catte16, nmrlipids_proj4}
and to mark the improvement after treating electronic polarizability, 
we show simulation results from ECC-lipids and also from Lipid17 \citep{lipid17-future}. 
While the response of the order parameters of PC to \ce{CaCl2} concentrations
is qualitatively correct in the Lipid14/17 model \cite{lipid17},
it is quantitatively comparable to the experimental trends in ECC-lipids. 
Moreover, the structure of PS head group is fundamentaly improved
after accounting for the electronic polarization 
(Fig.~\ref{fig:delta_ordPar_CaCl_PCPS}).

@@

The structural changes in POPS induced by increasing concentrations of \ce{CaCl2} 
are very challenging to interpret from both experiments \cite{roux90} 
and also from MD simulations \cite{nmrlipids_proj4}. 
The changes of the order parameters $S^\alpha$ and $S^\beta$ in PS head groups from simulations 
are not found to follow the experimental profiles, 
nor any common trend,
with the currently available models \cite{nmrlipids_proj4}. 
In Fig.~\ref{fig:delta_ordPar_CaCl_PCPS},
we plot these structural parameters for one of the models in that work, Lipid17, 
and for ECC-lipids. 
Although the agreement between ECC-lipids and experiments is far from optimal,
the general trends and behaviour even at highly concentrated solutions of \ce{CaCl2} 
are well captured by the model. 
For example the order parameter $\beta$ in POPS 
adopts a saturated value in experiments and ECC-lipids,
but it is growing with the \ce{CaCl2} concentration in Lipid17.
In addition, 
the forking of the order parameters $\alpha$ 
is almost constant in the experiment and ECC-lipids,
but it varies greatly in th Lipid17.

@@

Such a largely diminished flexibility 
may arise from the binding of calcium to the \ce{COO^-} group in PS. 
Whe have measured the mean orientation of the \ce{COO^-} group 
as the connector of the carbon atoms 
forming the bond between the group and the $\beta$-carbon of the phospholipid. 
The mean angle between the \ce{COO^-} vector and the membrane normal
changes by $11^\circ$ from $62^\circ$ to $73^\circ$ (more towards the membrane interior)
 at 620~mM buffer concentration of \ce{CaCl2}.
The conformational flexibility of the PS head group is hence restrained 
by the binding of calcium cations to the \ce{COO^-} group, 
which reorients towards the membrane interior and to the phosphate region, 
where calcium cations bind predominantly.

@@

The binding of calcium cations to the phospholipid bilayer is shown 
in the density profiles of \ce{Ca^{2+}}, \ce{Na^+} counterions 
and also \ce{Cl^-} ions in Fig.~\ref{fig:cacl-dens_PCPS}. 
Clearly, after accounting for the electronic polarizability, 
the population of the cations in the phosphate region is significantly diminished. 
\todo{JOE: The density plot still needs to be changed to show both Lipid17 and ECC-lipids, and to support this argument.}
Still, the dominant contribution to the binding of calcium comes from the phosphate groups. 
Interestingly enough,
although the \ce{COO^-} groups increase the affinity of calcium towards PS compared to PC,
 there is no extra density at the outmost surface of the membrane. 
This can be understood from the point of the structural rearangements of the \ce{COO^-} group, 
which take place during the interactions with calcium cations
described above and in Fig.~\ref{fig:delta_ordPar_CaCl_PCPS}.

The higher affinity of calcium to PS head groups compared to PC
is well demonstrated using the relative surface excess, $\Gamma ^w _{Ca}$,
summarized in Table~\ref{tab:binding}. 
While $\Gamma^{w}_{Ca}$ at a neutral POPC bilayer
is       $\Gamma^{w}_{Ca} = 0.06\mathrm{nm^{-2}}$, 
its value for the PC:PS (5:1) bilayer is $\Gamma^{w}_{Ca} = 0.24\mathrm{nm^{-2}}$
at comparable buffer concentrations  $350 \, \mathrm{mM}$ resp. $400\, \mathrm{mM}$.

@@

Further details about the interactions of \ce{Ca^{2+}} with various moieties in POPC or POPS
were obtained by counting contacts between the cations and the oxygen atoms of the lipids
similarly as was done in \cite{melcr18}. 
The threshold for counting a contact was set to $0.3\,\mathrm{nm}$, 
which encompasses the first peak of the radial distribution function between the cations and the oxygen atoms of the lipids.

The percentages of the populations of membrane-bound calcium cations for various membrane moieties 
are summarized in Table~\ref{tab:binding}.
Even though the negatively charged membrane contains only 18\% of POPS, 
approximately 41\% of the total population of bound calcium cations is in contact with PS lipids
with 8\% bound only to them. 
This corroborates the intrinsically higher affinity of PS lipids to calcium cations compared to neutral PC lipids. 
POPC interacts with the calcium cations almost entirely through its phosphate group 
in both neutral and negatively charged membranes. 
Interactions of \ce{Ca^{2+}} with carbonyl groups are also present, 
however, they are always accompanied by interactions with phosphate groups. 
%but a very small negligable amount much below 1\%. 
\todo{Also COO- vs. phosphate binding should be discussed and the results should be
  discussed with respect to the literature: https://doi.org/10.1016/j.bpj.2018.08.044 and
  Melcrova et al. Scientific Reports | 6:38035 | DOI: 10.1038/srep38035 }
%
\todo{EDIT: Taken from NMRlipids4 as an interesting information, to be edited.}
The dehydrated complexes of PS headgroup and calcium ions can also lead to the
phase separation \cite{hauser77,kurland79,hauser85,feigenson86,mattai89,roux90,roux91}.
%
CHARMM36 force field \cite{klauda10,venable13} combined with 2D infrared spectroscopy
suggests that calcium ions interacts only with the carboxylate group of PS lipids \cite{valentine18}, while
the same force field without the NBfix parameters together with the NMR chemical shifts and
REDOR experiments suggests a significant binding affinity also to the phosphate region \cite{hallock18}.
On the other hand, simulations with the Berger force field \cite{berger97,mukhopadhyay04}
combined with fluorescent and vibrational sum frequency spectroscopy suggested a significant
calcium binding also to the carbonyls in the acyl chains \cite{melcrova16}.

@@

Relative probabilities of \ce{Ca^{2+}} complexes with a certain number of lipids are presented in Fig.~\ref{fig:cacl_complexes}. 
Calcium cations that are bound only to PC in the mixed bilayer with PS 
behave similarly as in the pure PC bilayer
maintaining comparable probabilities for clustering one, two, or even three PC lipids together. 
In contrast, PS lipids prefer 1:1 ratio with \ce{Ca^{2+}},
which may also be due to their low molar fraction in the the mixed bilayer. 
In total, however, the negatively charged membrane has its stoichiometry 
shifted towards complexes with three phospholipids and one calcium. 
This is obviously due to the presence of POPS, 
which also contributes to the pre-formed one- and two-membered clusters of POPC. 
Clusters of four or more lipids were not observed in either membrane.

%This is also reflected in the increased probability 
%of a single phospholipid interacting with two \ce{Ca^{2+}} cations 
%compared to the neutral POPC bilayer, 
%where such a probability is almost negligable. 
% Not shown through numbers, however, I have the plots to back it.

%\todo{Stationary distribution: Make a figure documenting the populations of bound \ce{Ca^{2+}} cations (like I have in the presentation) that would accompany Table \ref{tab:Ca_binding_PCPS}. 
%This will roughly correspond to the PC stoichiometry plot \ref{fig:cacl_complexes}. }

@@

%In addition to such estimates of the time scales, we used the Markov model on top of the simulation with the negatviely charged mixed bilayer
%to calculate the time of the mean first passage of calcium from solution to the membrane (and in reverse) resulting in $55\,\mathrm{ns}$  ($165\,\mathrm{ns}$).

%\todo{Fluxes: committor analysis, dominant fluxes (table and figure)}
%From the spectrum of the transition matrix, we also observe that 
%the slowest transitions are associated with the binding to two or three phosphate moieties from either PC or PS,
%which are also among the states with the highest probabilities.
%The net fluxes of calcium cations from solution to such states also form a large contribution to the total flux ($\approx 45\%$). 
%\todo{Make a table and a figure of the state probabilities and fluxes to support this statement.}. 
%Analysis of the possible binding pathways of calcium cations to the negatively charged mixed bilayer 
%reveal that the cations mostly enter the membrane bound states directly from solution 
%without complicated transitions at the time scales of the order of the lag time of the Markov state model, $25\,\mathrm{ns}$.

Provide density plots and show, where and how the cations bind to the membrane.

\todo{SAMULI: I have putted here two preliminary sentences}
Applying ECC to POPS lipids improves cation binding affinity to membranes
mixed with POPC and also the response of PS headgroup to the bound cations.
However, further improvement of the parameters is needed to fully reproduce the
details of calcium binding to PS headgroups.

>>>>

The interaction of various cations including calcium 
with  neutral and also negatively charged lipid bilayers
was thoroughly compared between simulations and experiments 
from the point of the electrometer concept \cite{roux90, seelig87, altenbach84}
in the references~\citenum{catte16, nmrlipids_proj4}. 
It was concluded that the interaction of calcium with zwitterionic PC bilayers 
is generally overestimated in all studied models \cite{catte16},
while none of the models was capable of interpreting the NMR structural parameters
with the negatively charged PC:PS (5:1) bilayers \cite{nmrlipids_proj4}. 
Accounting for the electronic polarization was identified as 
a non-negligible correction to the interaction between cations and POPC bilayers in Ref.~\cite{melcr18}.

Here, we demonstrate that the same approach 
significantly improves the description of calcium binding 
also to the negatively charged PC:PS (5:1) bilayers. 
In correspondence with the Ref.~\citenum{nmrlipids_proj4},
we also use the changes of the head group order parameters $S^\alpha$ and $S^\beta$ 
in PC and PS lipids 
to quantify the binding affinity of the calcium cations
and the induced structural changes
(Fig.~\ref{fig:delta_ordPar_CaCl_PCPS}). 
To directly connect to the previous works \citep{catte16, nmrlipids_proj4}
and to mark the improvement after treating electronic polarizability, 
we show simulation results from the newly developed ECC-lipids and also from Lipid17 \citep{lipid17-future}. 
While the response of the order parameters of PC to \ce{CaCl2} concentrations
is qualitatively correct in the Lipid14/17 model \cite{lipid17},
it is quantitatively comparable to the experimental trends in ECC-lipids. 
Moreover, the structure of PS head group is fundamentaly improved
after accounting for the electronic polarization 
(Fig.~\ref{fig:delta_ordPar_CaCl_PCPS}).

@@

The structural changes in POPS induced by increasing concentrations of \ce{CaCl2} 
are very challenging to interpret from both experiments \cite{roux90} 
and also from MD simulations \cite{nmrlipids_proj4}. 
The changes of the order parameters $S^\alpha$ and $S^\beta$ in PS head groups from simulations 
are not found to follow the experimental profiles, 
nor any common trend,
with the currently available models \cite{nmrlipids_proj4}. 
In Fig.~\ref{fig:delta_ordPar_CaCl_PCPS},
we plot these structural parameters for one of the models in that work, Lipid17, 
and for ECC-lipids. 
Although the agreement between ECC-lipids and experiments is not optimal,
the general trends and behaviour even at highly concentrated solutions of \ce{CaCl2} 
are well captured by the model. 
For example the order parameter $\beta$ in POPS 
adopts a saturated value in experiments and ECC-lipids,
but it is growing with the \ce{CaCl2} concentration in Lipid17.
In addition, 
the forking of the order parameters $\alpha$ 
is almost constant in the experiment and ECC-lipids,
but it varies greatly in th Lipid17.

@@

Such a largely diminished flexibility 
may arise from the binding of calcium to the \ce{COO^-} group in PS. 
Whe have measured the mean orientation of the \ce{COO^-} group 
as the connector of the carbon atoms 
forming the bond between the group and the $\beta$-carbon of the phospholipid 
(Fig.~\ref{fig:delta_ordPar_CaCl_PCPS}).
The mean angle between the \ce{COO^-} vector and the membrane normal
changes by $11^\circ$ from $62^\circ$ to $73^\circ$ (more towards the membrane interior)
 at 620~mM buffer concentration of \ce{CaCl2}.
The conformational flexibility of the PS head group is hence restrained 
by the binding of calcium cations to the \ce{COO^-} group, 
which reorients towards the membrane interior and to the phosphate region, 
where calcium cations bind predominantly.

@@

The binding of calcium cations to the phospholipid bilayer is shown 
in the density profiles of \ce{Ca^{2+}}, \ce{Na^+} counterions 
and also \ce{Cl^-} ions in Fig.~\ref{fig:cacl-dens_PCPS}. 
Clearly, after accounting for the electronic polarizability, 
the population of the cations at the membrane interface is significantly diminished. 
\todo{JOE: The density plot still needs to be changed to show both Lipid17 and ECC-lipids, and to support this argument.}
Still, the dominant contribution to the binding of calcium comes from the phosphate groups. 
Interestingly enough,
although the \ce{COO^-} groups increase the affinity of calcium towards PS compared to PC,
 there is no extra density at the outmost surface of the membrane. 
This can be understood from the point of the structural rearangements of the \ce{COO^-} group, 
which take place during the interactions with calcium cations
described above and in Fig.~\ref{fig:delta_ordPar_CaCl_PCPS}.

\todo{JOE: suggestion -- what is the charge at the membrane interface of memb+counterions -vs- memb+CaCl2 concnetration?
                         this could relate well to the zeta potential measurements and the Fig 1 in \cite{melcrova16}. }

@@

% Gamma and higher affinity of PS to CA2+
The higher affinity of calcium to PS head groups compared to PC
is well demonstrated using the relative surface excess, $\Gamma ^w _{Ca}$,
summarized in Table~\ref{tab:binding}. 
While $\Gamma^{w}_{Ca}$ at a neutral POPC bilayer
is       $\Gamma^{w}_{Ca} = 0.06\mathrm{nm^{-2}}$      \cite{melcr18}, 
its value for the PC:PS (5:1) bilayer is $\Gamma^{w}_{Ca} = 0.24\mathrm{nm^{-2}}$
at comparable buffer concentrations  $350 \, \mathrm{mM}$ resp. $400\, \mathrm{mM}$.

% Description of no. conctacts counting
Such a significantly increased binding also has an impact 
on the structure and populations of the bound calcium cations. 
We have counted contacts between the cations and the oxygen atoms of the lipids
similarly as was done in \cite{melcr18} 
to obtain such a detailed information 
about the interactions of \ce{Ca^{2+}} with various moieties in POPC or POPS. 
The threshold for counting a contact was set to $0.3\,\mathrm{nm}$, 
which encompasses the first peak of the radial distribution function between the cations and the oxygen atoms of the lipids.

% Populations from contacts
The percentages of the populations of membrane-bound calcium cations for various membrane moieties 
are summarized in Table~\ref{tab:binding}.
Even though the negatively charged membrane contains only 18\% of POPS, 
on average 41\% of the total population of bound calcium cations is in contact with PS lipids
with 8\% bound only to them. 
% Where Calcium interacts
While POPC interacts with the calcium cations almost entirely through its phosphate group 
in both neutral \cite{melcr18} and negatively charged membranes, 
the inteaction with POPS lipids happens more through the \ce{COO^-} group than through the \ce{PO4} group. 
Interactions of \ce{Ca^{2+}} with carbonyl groups are also present for both POPC and POPS,
however, they are always accompanied by interactions with phosphate groups. 
This is in contrast to 
simulations with the Berger force field \cite{berger97,mukhopadhyay04}
combined with fluorescent and vibrational sum frequency spectroscopy,
which suggest a significant
calcium binding also to the carbonyls groups \cite{melcrova16}. 
This feature arises likely from the inaccuracy of the Berger model 
and its overestimated binding of cations to that region \cite{nmrlipids_proj4}.

Measurements using infrared spectroscopy 
combined with simulations with the CHARMM36 model 
with a correction for overestimated binding of calcium \cite{valentine18}
reveal similar trends for the populations at various PS moieties
suggesting that POPS interacts mostly through its \ce{COO^-} group. 
While this feature agrees nicely with our findings, 
the balance of the interactions of PS moieties with calcium cations may be suboptimal
and the interaction strength underestimated in total in those simulations \cite{nmrlipids_proj4}.

In contrast, the combination of NMR experiments 
and MD simulations with the same model, CHARMM36, but without the correction for calcium binding
also reveal long-lived interactions of \ce{Ca^{2+}} with \ce{PO4} groups. \cite{hallock18}
Although the ratios of the life times of coordination follow the same trends as in our work,
the lengths of simulation trajectories are insufficient 
to yield converged results in this aspect (see SI and Refs.~\cite{catte16, melcr18, nmrlipids_proj4}). 
Moreover, the content of POPS in the bilayers in that work is
much higher above the threshold after which 
\ce{Ca^{2+}} cations induce lipid clustering. \cite{hauser77,kurland79,hauser85,feigenson86,mattai89,roux90,roux91}
At such a high content of negatively charged lipids,
PS head groups are found to form even four-membered clusters \cite{hallock18}. 
Our simulations are, however, at different conditions,
in which the clustering of POPS does not happen, 
and PS lipids prefer 1:1 ratio with \ce{Ca^{2+}}
as plotted in Fig.~\ref{fig:cacl_complexes}
showing relative probabilities of \ce{Ca^{2+}} complexes with a certain number of PC or PS lipids. 
Compared to a neutral POPC bilayer
the negatively charged bilayer has its stoichiometry 
shifted towards complexes with three phospholipids and one calcium. 
This is obviously due to the presence of POPS, 
which also contributes to the pre-formed one- and two-membered clusters of POPC. 
Clusters of four or more lipids were not observed in either membrane.

\todo{ THE REMAINING PIECES TO BE DISCUSSED  --  consider wheter the dynamics slowdown or the dihedral distributions could make good contributions to this paper. }

% Ca2+ induces much slower dynamics of lipids when PS is present,
% this does not happen with PC
% uses NBFIX
From Ref.~\cite{valentine18}://
To summarize, calcium binding to PS lipids slows the dynamics at the ester position and introduces more heterogeneous molecular environments. 
The slowdown cannot be attributed to complete ester dehydration or decreased hydrogen bonding to water, so we interpret these results as evidence that the lipid headgroups adopt more rigid but disordered conformations on the molecular level. 
The decrease in dynamics observed experimentally indicates that the network of hydrogen bonds between lipids and water at the polar-nonpolar interface becomes slower, as it was shown by Karathanou and Bondar (34). 
This is bolstered by recent NMR work showing that binding to Ca2+ induces two different rigid conformations in the glycerol backbone of lipid headgroups (18).
We have investigated the molecular origin of the rigidification and decreased lateral diffusion, finding that calcium binding does not dehydrate the ester region, but even with only 10\% PS lipids, the local dynamics slow around the esters. The slower local dynamics provide a molecular explanation for the Ca2+-dependent decreases in fluidity observed on longer length and timescales.
There is considerable room for further improvement in the force fields for ion-membrane interactions. Particularly, small differences in short-range interactions may determine the local geometries (34
). Most of our conclusions are, however, based on long-range effects (i.e., electric field generated by the ion at significant distances from the carbonyl groups), which are less sensitive to the details of the force field.

Their Figure 5 \cite{hallock18} shows the distributions of O-Ca-Cb-N dihedral angle
revealing separate populations of the PS head groups. 
These two structures are also compared in therm of P-Cg distances, 
which can be measured from NMR REDOR exp. 
It is concluded that since the two distinct populations in P-Cg distance can be represented with the 
dihedral distributions, 
they also likely represent the distinct populations. 
 -> plotting the distribution of O-Ca-Cb-N dihedral angle may be also a good idea for us (SI?)
     -> NMRlipids 4 in SI \cite{nmrlipids_proj4} plots this information: 
           - C36 agress with the distribution in \cite{hallock18}
           - Lipid17 (and Slipids) are almost identical to C36
           - but the structure of the COO- is highly different (dih angle Ca-Cb-Cg-Og)

@@

\todo{SAMULI: I have putted here two preliminary sentences
JOSEF: I can write conclusions, but I would prefer to wait with that after your comments and edits. }
Applying ECC to POPS lipids improves cation binding affinity to membranes
mixed with POPC and also the response of PS headgroup to the bound cations.
However, further improvement of the parameters is needed to fully reproduce the
details of calcium binding to PS headgroups.

\section{SI}

@@
